mathml: Translate R expressions to MathML and LaTeX

### 2023-01-22 rj/mathml.Rmd
customizing

Schwarz [-@schwarz1994] applied the model to data from a redundant signals task
with 13 onset
asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
where \(0\) refers to the synchronous condition, and \(\pm\infty\) to the
single-target presentations. Each condition was replicated 400 times. The
observed mean response times and their standard deviations are given in
Table \@ref(tab:miller-data) below.

@@

>>>>

Schwarz [-@schwarz1994] applied the model to data from a redundant signals task
with 13 onset
asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
where \(\tau = 0\) refers to the synchronous condition, and \(\pm\infty\) to the
single-target presentations. Each condition was replicated 400 times. The
observed mean response times and their standard deviations are given in
Table \@ref(tab:miller-data) below.

@@

# Appendix: Customizing the package

Several ways exist for translating new R\ terms to their mathematical
representation. We have already seen above how to use "hooks" to translate long
variable names from R to compact mathematical signs, as well as functions such
as cumulative probabilities \(P(X \le k)\) to different representations
like \(\sum_{i=0}^k P(X = i)\). Obviously, the hooks require that there already
exists a rule to translate the target representation into MathML and MathJax.

In this appendix we describe a few more ways to extend the set of translations
according to a user's needs. As stated in the background section, the Prolog end
provides two classes of rules for translation, macros `math/2,3,4` mirroring the
R\ hooks mentioned above, and the low-level predicates `ml/3` and `jax/3` that
create proper MathML and Latex terms.

## \(n\)-th root

Base R does not provide a function like `cuberoot(x)` or `nthroot(x, n)`, and
the present package does not support the respective representation. To obtain a
cube root, a programmer would typically type `x^(1/3)` or better `x^{1/3}` (see
the practice section why the curly brace is preferred in an exponent), resulting
in \(x^{1/3}\) which may still not match everyone's taste. Here describe the
steps needed to represent the \(n\)-th root as \(\sqrt[n]x\).

We assume that `nthroot(x, n)` is available in the current
namespace [manually defined, or from R package \pkg{pracma}, @pracma], so that
the names of the arguments and their order are accessible to `canonical()` if
needed. As we can see below, \pkg{mathml} uses a default
representation `name(arguments)` for such unknown functions.

```{r, results='asis'}
nthroot <- function(x, n)
  x^{1L/n}

term <- canonical(quote(nthroot(n=3L, 2L)))
mathout(term)
```

A proper MathML term is obtained by `mlx/3` (the x in mlx indicates that it is
an extension and is prioritized over the default ml/3 rules). `mlx/3`
recursively invokes `ml/3` for translating the function arguments _X_ and _N_,
and then contructs the correct MathML entity `<mroot>...</mroot>`.

```prolog
mlx(nthroot(X, N), M, Flags) :-
    ml(X, X1, Flags),
    ml(N, N1, Flags),
    M = mroot([X1, N1]).
```

The explicit unification `M = ...` in the last line serves to avoid clutter in
the head of `mlx/3`. The Prolog file `nthroot.pl` also includes the respective
rule for Latex and can be consulted from the package folder via the underlying
package \pkg{rolog}.

```{r, results='asis'}
rolog::consult(system.file(file.path("pl", "nthroot.pl"), package="mathml"))

term <- quote(nthroot(a * (b + c), 3L)^2L)
mathout(term)
term <- quote(a^(1L/3L) + a^{1L/3L} + a^(1.0/3L))
mathout(term)
```

The file `nthroot.pl` includes three more statements `precx/3` and `parenx/3`,
as well as a `math_hook/2` macro. The first sets the operator precedence of the
cubic root above the power, thereby putting a parentheses around nthroot
in \((\sqrt[3]{\ldots})^2\). The second tells the system to increase the counter
of the parentheses below the root, such that the outer parenthesis becomes a
square bracket.

The last rule maps powers like `a^(1L/3L)` to `nthroot/3`, as shown in the
first summand. Of course, \pkg{mathml} is not a proper computer algebra system.
As is illustrated by the other terms in the sum, such macros are limited to
purely syntactical matching, and terms like `a^{1L/3L}` with the curly brace
or `a^(1.0/3L)` with a floating point number in the numerator are not detected.
